---
title: "Week 1, Day 1: Setup + R Basics (self-guided)"
subtitle: "Foundations of Data Science with R"
author: "Your Name"
date: "`r Sys.Date()`"
output: github_document
---

### Before you start

We do **not** meet as a class on Monday 08/31 (Convocation). Instead, work through this
page on your own before our first class meeting on **Wednesday 09/02**.
Budget **60–90 minutes**, most of it for installing software.

If you get stuck on setup, don't burn hours on it alone — come to office
hours or email me (mgelman@cmc.edu) and we'll sort it out together
before Wednesday.

------------------------------------------------------------------------

### What is data science, and why R?

Before you install anything, a quick orientation on what this course is
actually about.

**Data science** is the practice of using data to answer questions and
make decisions. Most projects move through a similar cycle: **ask** a
question, **wrangle** (get and clean) the data, **explore** it visually
and numerically, **model** it, and **communicate** what you found. This
course is structured around that cycle — Part I of the semester covers
wrangling and exploration, Part II covers modeling.

We'll do all of this in **R**, a programming language built specifically
for data and statistics. A few reasons it's a good fit:

-   A huge ecosystem of packages for nearly any data task (the
    "tidyverse," which you'll use constantly).
-   `ggplot2`, one of the most capable data visualization tools available
    in any language.
-   It was built by statisticians, so the statistical methods you'll
    learn later this semester (regression, classification, etc.) are
    first-class, not an afterthought.
-   Strong tools for reproducibility — R Markdown (which you'll use for
    every assignment, including this one) combines code, output, and
    narrative into one document that anyone can re-run and get your exact
    results.

(Python is also widely used for data science and you'll likely run into
it elsewhere — R is simply the tool this course uses.)

That's the "why." The rest of this page is the "how": getting your tools
installed so you're ready to work in R starting next class.

------------------------------------------------------------------------

### Setup: R, RStudio, Git, GitHub

Follow the full setup guide here, start to finish:

**[github.com/econ122-f26/github-classroom-for-students](https://github.com/econ122-f26/github-classroom-for-students)**

It walks you through installing R and RStudio, installing Git, creating a
GitHub account, configuring Git, and setting up a Personal Access Token —
everything you need connected and working together. Come back to this
page once you can knit an R Markdown document (the guide will get you
there).

------------------------------------------------------------------------

### A note on R Markdown: running vs. knitting

This document is an R Markdown file (`.Rmd`). It's important to
understand the difference between running code chunks and knitting the
document.

-   **Running a Code Chunk:** Click the green "play" button, or press
    `Ctrl+Enter` (`Cmd+Enter` on Mac), to execute code directly in the
    R console. Useful for testing as you go.
-   **Knitting the Document:** Click the **Knit** button to run every
    code chunk top to bottom in a clean session, producing a final,
    reproducible output file. This is how you'll submit assignments —
    it proves your analysis runs start to finish without errors.

Try it now: open this file in RStudio and click **Knit**.

------------------------------------------------------------------------

### Part 1: R Basics and Variable Assignment (10 minutes)

**Task 1.1: Hello, R!**
Use R as a calculator to find the result of $$(150 + 75) / 3$$ in the code chunk below.

```{r}
# Write your code here to calculate the result
```

**Task 1.2: Creating Variables**
Create three new variables in the code chunk below.
-   `my_age`: Your age as a numeric value.
-   `my_name`: Your name as a character string.
-   `is_student`: A logical value (`TRUE` or `FALSE`) to indicate if you are a student.

```{r}
# Write your code to create the variables here
```

**Task 1.3: Combining Variables**
Use the `paste()` function to create a new variable called `greeting` that combines your name and a welcome message. For example: "Hello, `Your Name`! Welcome to Data Science."

```{r}
# Write your code to create the greeting variable here
```

### Part 2: Exploring Data Types & Logical Operations (15 minutes)

**Task 2.1: Checking the Class**
Use the `class()` function to check the data type of the three variables you created in Part 1. Explain what each output means in the space below.

```{r}
# Write your code to check the class of each variable here
```

**Task 2.2: Integer vs. Numeric**
Create two variables, one storing the number 10 as a numeric and another storing 10 as an integer. Use the `class()` function to show the difference.

*Hint: Remember to use the `L` suffix for integers.*

```{r}
# Write your code here to demonstrate the difference
```

**Task 2.3: Logical Operations**
Use logical operators (`>`, `<`, `==`, `!=`) to answer the following questions. Write the R code and the output you expect.

1. Is $10$ greater than $5$?
2. Is "R" the same as "r"?
3. Is the variable `is_student` true?

```{r}
# Write your code to check the logical conditions here
```

### Part 3: Introduction to Vectors (15 minutes)

**Task 3.1: Creating a Vector**
Vectors are the most basic data structure in R. Use the `c()` function to create a vector named `my_numbers` with the values $5, 10, 15, 20$. Then, print the vector.

```{r}
# Create and print the vector here
```

**Task 3.2: Vector Operations**
Add a new number, $25$, to your `my_numbers` vector. You can do this by re-assigning the variable. Then, calculate the sum and the mean of the new vector.

```{r}
# Write your code here to update the vector and perform calculations
```

**Task 3.3: Logical Vectors**
Create a new vector called `is_even` that contains logical values (`TRUE` or `FALSE`) for each element in your `my_numbers` vector, indicating whether the number is even.

*Hint: The modulo operator `%%` is useful for this.*

```{r}
# Write your code here to create the logical vector
```

### Challenge Question (for early finishers)

Imagine you have a vector of temperatures in Celsius: `c(10, 15, 20, 25)`. How would you convert this entire vector to Fahrenheit using a single line of R code? The formula for Celsius to Fahrenheit is

$$
(C \times 9/5) + 32
$$

.

```{r}
# Write the challenge question solution here
```

------------------------------------------------------------------------

### Before Wednesday

-   Confirm R, RStudio, and Git are installed and working.
-   Confirm you can knit this file to a `.md` or `.html` output.
-   Bring your laptop, fully set up, to class on 09/02.
