---
title: "Week 1, Day 1: Setup + R Basics (self-guided)"
subtitle: "Foundations of Data Science with R"
author: "Your Name"
date: "`r Sys.Date()`"
output: github_document
---

### Before you start

We do **not** meet as a class on Monday 08/31. Instead, work through this
page on your own before our first class meeting on **Wednesday 09/03**.
Budget **60–90 minutes**, most of it for installing software.

If you get stuck on setup, don't burn hours on it alone — come to office
hours or email me (mgelman@cmc.edu) and we'll sort it out together
before Wednesday.

------------------------------------------------------------------------

### Part 0: Install R and RStudio

1.  Install **R** from CRAN: <https://cran.r-project.org/>
    (choose your OS, take the latest release).
2.  Install **RStudio Desktop** (free version):
    <https://posit.co/download/rstudio-desktop/>
3.  Open RStudio and confirm it launches. You should see four panes:
    Console, Source, Environment/History, and Files/Plots/Packages/Help.

### Part 0.5: Install Git and create a GitHub account

We'll use Git and GitHub for every assignment this semester, starting
next week, so it's worth getting comfortable now.

1.  Create a free account at <https://github.com/> if you don't have one.
    Use a professional username — you may end up linking this account to
    a resume or portfolio.
2.  Install Git:
    -   **Windows:** <https://gitforwindows.org/>
    -   **Mac:** open Terminal and run `git --version` — this usually
        prompts you to install Xcode Command Line Tools, which include Git.
3.  Follow **Chapters 4–9** of *Happy Git and GitHub for the useR*
    (<https://happygitwithr.com/>) to connect Git, GitHub, and RStudio.
    This is the single best resource for getting this stack working —
    don't skip it if `usethis::git_sitrep()` (covered in the guide)
    reports problems.

### Part 0.75: Join GitHub Classroom

*(Link TBD — I'll post the GitHub Classroom invite link here and announce
it separately before it's needed. You do not need this for today's
activity.)*

------------------------------------------------------------------------

### A note on R Markdown: running vs. knitting

This document is an R Markdown file (`.Rmd`). It's important to
understand the difference between running code chunks and knitting the
document.

-   **Running a Code Chunk:** Click the green "play" button, or press
    `Ctrl+Enter` (`Cmd+Enter` on Mac), to execute code directly in the
    R console. Useful for testing as you go.
-   **Knitting the Document:** Click the **Knit** button to run every
    code chunk top to bottom in a clean session, producing a final,
    reproducible output file. This is how you'll submit assignments —
    it proves your analysis runs start to finish without errors.

Try it now: open this file in RStudio and click **Knit**.

------------------------------------------------------------------------

### Part 1: R Basics and Variable Assignment (10 minutes)

**Task 1.1: Hello, R!**
Use R as a calculator to find the result of $$(150 + 75) / 3$$ in the code chunk below.

```{r}
# Write your code here to calculate the result
```

**Task 1.2: Creating Variables**
Create three new variables in the code chunk below.
-   `my_age`: Your age as a numeric value.
-   `my_name`: Your name as a character string.
-   `is_student`: A logical value (`TRUE` or `FALSE`) to indicate if you are a student.

```{r}
# Write your code to create the variables here
```

**Task 1.3: Combining Variables**
Use the `paste()` function to create a new variable called `greeting` that combines your name and a welcome message. For example: "Hello, `Your Name`! Welcome to Data Science."

```{r}
# Write your code to create the greeting variable here
```

### Part 2: Exploring Data Types & Logical Operations (15 minutes)

**Task 2.1: Checking the Class**
Use the `class()` function to check the data type of the three variables you created in Part 1. Explain what each output means in the space below.

```{r}
# Write your code to check the class of each variable here
```

**Task 2.2: Integer vs. Numeric**
Create two variables, one storing the number 10 as a numeric and another storing 10 as an integer. Use the `class()` function to show the difference.

*Hint: Remember to use the `L` suffix for integers.*

```{r}
# Write your code here to demonstrate the difference
```

**Task 2.3: Logical Operations**
Use logical operators (`>`, `<`, `==`, `!=`) to answer the following questions. Write the R code and the output you expect.

1. Is $10$ greater than $5$?
2. Is "R" the same as "r"?
3. Is the variable `is_student` true?

```{r}
# Write your code to check the logical conditions here
```

### Part 3: Introduction to Vectors (15 minutes)

**Task 3.1: Creating a Vector**
Vectors are the most basic data structure in R. Use the `c()` function to create a vector named `my_numbers` with the values $5, 10, 15, 20$. Then, print the vector.

```{r}
# Create and print the vector here
```

**Task 3.2: Vector Operations**
Add a new number, $25$, to your `my_numbers` vector. You can do this by re-assigning the variable. Then, calculate the sum and the mean of the new vector.

```{r}
# Write your code here to update the vector and perform calculations
```

**Task 3.3: Logical Vectors**
Create a new vector called `is_even` that contains logical values (`TRUE` or `FALSE`) for each element in your `my_numbers` vector, indicating whether the number is even.

*Hint: The modulo operator `%%` is useful for this.*

```{r}
# Write your code here to create the logical vector
```

### Challenge Question (for early finishers)

Imagine you have a vector of temperatures in Celsius: `c(10, 15, 20, 25)`. How would you convert this entire vector to Fahrenheit using a single line of R code? The formula for Celsius to Fahrenheit is

$$
(C \times 9/5) + 32
$$

.

```{r}
# Write the challenge question solution here
```

------------------------------------------------------------------------

### Before Wednesday

-   Confirm R, RStudio, and Git are installed and working.
-   Confirm you can knit this file to a `.md` or `.html` output.
-   Bring your laptop, fully set up, to class on 09/03.
